Multi-Instance Methods for Partially Supervised Image Segmentation

### 2013-09-16 partial_supervision/paper.tex
Shorten some titles because they were ugly
trying to rescue semi supervised stuff

\chapter{Weakly Supervised Object segmentation via Multiple Instance Learning}

Most algorithms for semantic image segmentation and object-class segmentation
work with strong supervision: a pixel-wise labeling of training images. In this
chapter, we will investigate a method that works with annotation that is much
easier to obtain: whole image labels.

@@

We formulate weakly supervised multi-class image segmentation as a multi-instance problem, based on candidate segments.
In multi-instance learning~\citep{dietterich1997solving}, each training example is given as a multi-set of instances, called a bag.
Each instance is represented as a feature vector $x$ and a label $y$.
A bag is labeled positive if it contains at least one positive example, and negative otherwise.
During training, only the labels of the training bags, not of the instances inside the bags, are known.
The goal is to learn a classifier for unseen bags. 
Formally, let $\mathcal{X}$ be the set of instances. To simplify notation, we assume that bags are simply sets, not multi-sets.
Then a bag is an element of the power set $2^\mathcal{X}$ and the task is
to learn a function
\begin{equation} f_{MI} \colon 2^\mathcal{X} \rightarrow \{-1,+1\}.  \end{equation}
Training examples are tuples $(X_i,y_i)$ of bags $X_i \subset \mathcal{X}$ and labels $y_i \in \{-1,+1\}$.
It is often assumed that the $f_{MI}$ function stems from a so-called underlying concept, given by an (unknown) function
$f_{I} \colon \mathcal{X} \rightarrow \{-1,+1\}$, with 
\begin{equation}\eqlabel{multi_instance}
f_{MI}(X)= \max_{x \in X}\ f_{I}(x).
\end{equation}

Sometimes, the goal of finding $f_{MI}$ is extended to finding labels not only
for new bags, but also for all the separate instances within a bag
\citep{liconvex2010,zha2008joint}, that is finding $f_{I}$.\\ 
Even though finding $f_{I}$ is sometimes included in the task statement, there
has been very little work that actually reported accuracy on instance label
prediction. Part of the reason for this might be that for many of the datasets
used in multi-instance learning no ground truth for instances exists.

We look explicitly at accuracy on instance-level since we are interested in
actually segmenting images, not just classifying them. For multi-class image
segmentation, there are some hand-labeled datasets that provide ground truth
on pixel level. We use this ground truth to evaluate the performance of our
method. This approach does not exactly correspond to instance-level ground truth --
since the instances are segments, not pixels -- but relates to it closely.

%TODO theoretical part reference?
% TODO repetition below?

This chapter explores the 
application of multi-instance learning algorithms to the task of partially supervised image segmentation.
Multi-instance learning is a natural formulation for image classification and has been
successfully applied in this task~\citep{zhou2007multi}. We propose to go a step further and apply
multi-instance learning to the task of object-class segmentation in natural
images.
% FIXME relate to alex here
To our knowledge, all previous methods in the field use strong
supervision, meaning manual pixel-wise annotation of training images. This
approach does not scale to larger datasets, especially if one expects
consistency and quality in the segmentations.

After our empirical evaluation on image segmentation, we will extend the existing theoretical framework
on instance classification based on multi-instance prediction.

@@

\citet{gaertner2002multi} introduced the concept of a multi-instance
kernel on bags, defined in terms of a kernel on instances. 
The basic principle of multi-instance kernel is similar to a soft-max over instances in
each bag. This can be viewed as approximating the kernel value of the ``closest pair'' given by two bags. They show that the multi-instance kernel
is able to separate bags if and only if the original kernel on instances is able to separate the underlying concepts.
The method of \citet{gaertner2002multi} has a particular appeal in that it transforms a multi-instance problem into a standard
classification problem by using an appropriate kernel. The downside of this approach is that it does
not directly label instances, only bags.
 %and that it assumes that instances inside a bag have i.i.d.
%labels.

\citet{zhou2009multi} explicitly address the fact that instances are not
indpendent within a bag , leading to an algorithm that can take advantage of
possible correlations. Computational costs of their algorithm does not scale
well with the number of instances, although a heuristic algorithm is proposed
to overcome this restriction. 
%FIXME where does this sentence go?
\citet{zhou2009multi} demonstrated only a slight
advantage of their algorithm over the MI-kernel of \citet{gaertner2002multi},
so we use the MI-kernel for better scalability.

@@

\subsection{Semantic Scene Segmentation via Multi-Instance Learning}
Recently, several methods have been proposed to obtain semantic segmentations
of images using only image-level
supervision~\citep{vezhnevets2011weakly,verbeek2007region,vezhnevets2010towards}.
\citet{vezhnevets2011weakly}, for example, report impressive results on the
MSRC dataset.

While semantic segmentation is closely related to multi-class image
segmentation, there are important distinctions: In semantic segmentation, each
pixel has a semantic annotation, also containing non-object ``stuff'' classes like
``sky'', ``grass'' and ``water''. In multi-class image segmentation, the focus
is on objects, with possibly large parts of the image being labeled as
unspecific ``background''.  The unspecific background class contains much more
clutter than for example ``grass'' and is therefore much harder to model.
This makes disseminating the interesting part in multi-class object recognition
challenging, since it is not necessary possible to identify non-object regions
easily.

@@

\subsection{Multi-Instance Learning using MI-Kernels}
Since scalability is very important in real-world computer vision applications,
and natural images might need hundreds of segments to account for all possible
object boundaries, we use the efficient multi-instance
kernels~\citep{gaertner2002multi}.  Multi-instance kernels are a form of set
kernels %TODO add reference to earlier work
 that transform a kernel on instance level to a kernel on bag level.  We
reduce the multi-instance multi-class problem to a multi-instance problem by
using the one-vs-all approach.

@@

While using MIK has many advantages, it produces only an instance-level
classifier. We propose to transform a bag-level classifier $f_{MI}$ as given by
the SVM and \eqref{mi_kernel} into an instance-level classifier by setting
$f_{I}(x):=f_{MI}(\{x\})$, in other words, by considering each instance as its own
bag.

%TODO one classifier per class!

@@

The framework described above yields an image-level and a segment-level
classification.  In our setup, each segment might be given multiple labels.
%TODO explain!
To obtain a pixel-level object-class segmentation, we have to combine these.
When building the segmentation for a given image, we only consider classes
whose presence was predicted on image level.  Since we do not make use of the
ground truth segmentation during training, we cannot learn an optimal
combination as in \citet{li2010object} but perform a simple majority vote
instead.  We merge segments into pixel-level class labels by setting the label
$y_x$ of a pixel $x$ according to
\begin{equation}
    y_x = \argmax_{y \in Y} \#\{S_i | p \in S_i \land y_{S_i}=y \},
\end{equation}
where $Y= \{$car, bike, person$ \}$, $S_i$ enumerates all segments within an
%FIXME we didn't describe the dataset yet!
image and $y_{S_i}$ is the label of segment $S_i$. In words: each pixel is
assigned the class with the highest number of class segments containing it.
This simple heuristic yields good results in practice.

@@

We evaluate the performance of the proposed algorithm for object-class
segmentation on the challenging Graz-02 dataset.  This dataset contains 1096
images of three object classes, bike, car and person.  Each image may contain
multiple instances of the same class.

%From each image, we extract overlapping object-like segments using constraint
%parametric min cuts (CPMC,\citet{carreira2010constrained}). We use the top 200 ranked segments and discard the
%rest.
%Each segment is described using SIFT~\citet{lowe2004distinctive} and
%ColorSIFT~\citet{van2009evaluating} features, from which we compute bag of
%visual word histograms as well as histograms of oriented
%gradients~\citet{dalal2005histograms}. We construct one MI-Kernel per
%feature which are combined using multiple kernel learning.
We adjusted parameters on a hold-out validation set using bag-level information
and used the training and test sets as given by the dataset.  It is
straight-forward to extend the binary MIK method to the multi-class setting
using a one-vs-all strategy.
We train one multiple kernel learning (MKL) SVM per class using MIK and predict
class labels on segment level using MIK-instance. If at least one SVM
classifies a segment as positive, it is associated with the most confident
class. Otherwise, it is assigned ``background'' or no class.
This yields a classification of each segment into one of four classes: car,
bike, person, or background. We merge segments into pixel-level class labels as
described in \Secref{combiningsegments}.

@@

Per-class pixel accuracies are reported in \Tabref{tab:graz}; some qualitative
results are shown in \figref{results}. The overall accuracy on images labels,
which is the task that was actually trained, is $90\%$.  While the performance
of our multiple-instance based approach is far from current methods that use
pixel-level annotations, whose pixel-level accuracy is around
$70\%$~\citep{fulkerson2009class,schulz2011} on pixel-level, it can serve as a
baseline for research on weakly supervised methods for object class segmentation.

@@

We evaluated the proposed object-class segmentation method on the challenging
Graz02 dataset. While not reaching the performance of methods using strong
supervision, our result can work as a baseline for further research into weakly
supervised object class segmentation.

%In future work, we plan to scale our approach to much larger image datasets. As much
%more images with weak annotations are available than with pixel level segmentation,
%we hope that we can improve upon the state-of-the-art in object-class segmentation
%by making use of larger bodies of training images.

>>>>

\chapter{Weakly Supervised Object Segmentation}

Most algorithms for semantic image segmentation and object-class segmentation
work with strong supervision: a pixel-wise labeling of training images. In this
chapter, we will investigate a method that works with annotation that is much
easier to obtain: whole image labels.
While we do not reach the accuracy of competing fully supervised approaches,
our efficient, semi-supervised method is potentially able to scale to much
larger datasets, without the need of time-consuming manual annotation on
pixel-level.

@@

We formulate weakly supervised multi-class image segmentation as a
multi-instance problem, based on these candidate segments.  In multi-instance
learning~\citep{dietterich1997solving}, each training example is given as a
multi-set of instances, called a bag.  Each instance is represented as a
feature vector $x$ and a label $y$.

A bag is labeled positive if it contains at least one positive example, and
negative otherwise.  During training, only the labels of the training bags, not
of the instances inside the bags, are known.  The goal is to learn a classifier
for unseen bags. 
Formally, let $\mathcal{X}$ be the set of instances. To simplify notation, we
assume that bags are simply sets, not multi-sets.  Then a bag is an element of
the power set $2^\mathcal{X}$ and the task is to learn a function
\begin{equation} f_{MI} \colon 2^\mathcal{X} \rightarrow \{-1,+1\}.  \end{equation}
Training examples are tuples $(X_i,y_i)$ of bags $X_i \subset \mathcal{X}$ and
labels $y_i \in \{-1,+1\}$.  It is assumed that the $f_{MI}$ function stems
from a so-called underlying concept, given by an (unknown) function
$f_{I} \colon \mathcal{X} \rightarrow \{-1,+1\}$, with 
\begin{equation}\eqlabel{multi_instance}
f_{MI}(X)= \max_{x \in X}\ f_{I}(x).
\end{equation}

Multi-instance learning is a natural formulation for image classification and
has been successfully applied in this task~\citep{zhou2007multi}. We propose to
go a step further and apply multi-instance learning to the task of object-class
segmentation in natural images, by also classifying instances, not only bags.
In this, we follow the work of \citet{liconvex2010} and \citet{zha2008joint}, who not
only learn $f_{MI}$, but also $f_{I}$.

In our model, each image forms a bag, while the candidate segments correspond
to the instances contained in the bag. During learning, only presence of object
classes is needed as bag level supervision. By learning $f_{I}$, we are then
able to predict for individual segments whether they contain the object class
of interest, thereby obtaining a segmentation of the object.
To measure the performance of our algorithm, we use a dataset that not only contains
image level annotation, but also pixel level annotation of object. This allows
us to judge the succes of learning on instance level.

%This chapter explores the 
%application of multi-instance learning algorithms to the task of partially supervised image segmentation.
%FIXME not here!
% FIXME relate to alex here
%To our knowledge, all previous methods in the field use strong
%supervision, meaning manual pixel-wise annotation of training images. This
%approach is problematic, as scaling to larger datasets is limited by the
%effort to provide annotation, especially if one expects
%consistency and quality in the segmentations.

@@

\citet{gaertner2002multi} introduced the concept of a multi-instance
kernel on bags, defined in terms of a kernel on instances. 
The basic principle of the multi-instance kernel is similar to a soft-max over instances in
each bag. This can be viewed as approximating the kernel value of the ``closest pair'' given by two bags. \citet{gaertner2002multi} show that the multi-instance kernel
is able to separate bags if and only if the original kernel on instances is able to separate the underlying concepts.
The method of multi-instance kernels has a particular appeal in that it transforms a multi-instance problem into a standard
classification problem by using an appropriate kernel. The downside of this approach is that it does
not directly label instances, only bags.
 %and that it assumes that instances inside a bag have i.i.d.
%labels.

\citet{zhou2009multi} explicitly address the fact that instances are not
independent within a bag , leading to an algorithm that can take advantage of
possible correlations. Computational costs of their algorithm does not scale
well with the number of instances, although a heuristic algorithm is proposed
to overcome this restriction. 
%FIXME where does this sentence go?
\citet{zhou2009multi} demonstrated only a slight
advantage of their algorithm over the MI-kernel of \citet{gaertner2002multi},
so we use the MI-kernel for better scalability.

@@

\subsection{Semantic Scene Segmentation using Weak Annotation }
Recently, several methods have been proposed to obtain semantic segmentations
of images using only image-level
supervision~\citep{vezhnevets2011weakly,verbeek2007region,vezhnevets2010towards}.
\citet{vezhnevets2011weakly}, for example, report impressive results on the
MSRC dataset.

While semantic segmentation is closely related to multi-class image
segmentation, there are important distinctions: In semantic segmentation, each
pixel has a semantic annotation, also containing non-object ``stuff'' classes like
``sky'', ``grass'' and ``water''. In multi-class image segmentation, the focus
is on objects, with possibly large parts of the image being labeled as
unspecific ``background''.  The unspecific background class contains much more
clutter than for example ``grass'' and is therefore much harder to model. Additionally,
object classes themselves are much harder to capture using low-level textural information only.
This makes disseminating the distinctive features in multi-class object recognition
much more challenging.

@@

\subsection{Multi-Instance Learning using MI-Kernels}
Since scalability is very important in real-world computer vision applications,
and natural images might need hundreds of segments to account for all possible
object boundaries, we use the efficient multi-instance
kernels~\citep{gaertner2002multi}.  Multi-instance kernels are a form of set
kernels %TODO add reference to earlier work -> convolution kernel
 that transform a kernel on instance level to a kernel on bag level.  We
reduce the multi-instance multi-class problem to a multi-instance problem by
using the one-vs-all approach.

@@

While using MIK has many advantages, it produces only an instance-level
classifier. We propose to transform a bag-level classifier $f_{MI}$ as given by
the SVM and \eqref{mi_kernel} into an instance-level classifier by setting
$f_{I}(x):=f_{MI}(\{x\})$, in other words, by considering each instance as its own
bag. 
We extend the MIK SVM to the multi-class case using the one-vs-rest approach,
learning one SVM for each class.

@@

The framework described above yields an image-level and a segment-level
classification. In our setup, each segment might be given multiple labels.
To obtain a pixel-level object-class segmentation, we have to combine these.
As each image in the training set contains only a single class, we only consider the class
predicted by the bag-level one-vs-rest classifier.
Since we do not make use of the ground truth segmentation during training, we cannot learn an optimal
combination as in \citet{li2010object} but perform a simple majority vote
instead.  We merge segments into pixel-level class labels by setting the label
$y_x$ of a pixel $x$ according to
\begin{equation}
    y_x = \argmax_{y \in Y} \#\{S_i | p \in S_i \land y_{S_i}=y \},
\end{equation}
where $Y$ is the set of possible object classes, $S_i$ enumerates all segments within an
image and $y_{S_i}$ is the label of segment $S_i$. In words: each pixel is
assigned the class with the highest number of class segments containing it.
This simple heuristic yields good results in practice.

@@

We evaluate the performance of the proposed algorithm for object-class
segmentation on the challenging Graz-02 dataset.  This dataset contains 1096
images of three object classes, bike, car and person.  Each image may contain
multiple instances of the same class, but only one class is present per image.

We adjusted parameters on a hold-out validation set using bag-level information
and used the training and test sets as given by the dataset.  It is
straight-forward to extend the binary MIK method to the multi-class setting
using a one-vs-all strategy.
We train one multiple kernel learning (MKL) SVM per class using MIK and predict
class labels on segment level using MIK-instance. If at least one SVM
classifies a segment as positive, it is associated with the most confident
class. Otherwise, it is assigned ``background'' or no class.
This yields a classification of each segment into one of four classes: car,
bike, person, or background. We merge segments into pixel-level class labels as
described in \Secref{combiningsegments}.

@@

Per-class pixel accuracies are reported in \Tabref{tab:graz}; some qualitative
results are shown in \figref{results}. The overall accuracy on images labels,
which is the task that was actually trained, is $90\%$.  While the performance
of our multiple-instance based approach is far from current methods that use
pixel-level annotations, whose pixel-level accuracy is around
$70\%$~\citep{fulkerson2009class,schulz2011} on pixel-level, it can serve as a
baseline for research on weakly supervised methods for object-class segmentation.

@@

We evaluated the proposed object-class segmentation method on the challenging
Graz02 dataset. While not reaching the performance of methods using strong
supervision, our result can work as a baseline for further research into weakly
supervised object-class segmentation.
